\Title

### 2013-02-11 supplementary.tex
new Fig 4B plus some editing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Nonsynonymous changes outside of variable regions are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.5\linewidth]{synmut_conservation_4fold_synnonsyn}
\caption{Cumulative distribution of synonymous and nonsynonymous diversity in
{\it gag} in the LANL reference panel(filtered
sequences only, version 2011)~\cite{LANL2012}. Sites such that the consensus codon
has three synonymous and six nonsynonymous single mutants were used, and the number of observed mutants
of a certain type over the number of possible mutants is plotted. Nonsynonymous
changes are observed less often, {\it ergo} are more conserved, than
synonymous changes. It can be therefore assumed, as mentioned in the main text,
that non-escape nonsynonymous changes involve a large fitness cost.}
\label{fig:synnonsyncons}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Synonymous diversity across the HIV genome}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=\linewidth]{conservation_codons_genome}
\caption{Normalized codon entropy across the HIV genome is clustered around
variable regions in {\it env}. hence the synonymous mutational space is explored
deeper by HIV in that region, and even deleterious synonymous mutations can
reach high frequencies.
On the contrary, regions that are under purifying selection either
the the RNA level (the RRE hairpin) or at the protein level but in a different
reading frame (second {\it rev} exon, which includes the second {\it tat} exon),
show reduced entropy, i.e. more constrains.
All parts of {\it env} that are part of a different gene (signaling peptide,
second {\it rev} exon) have been excluded from our main analysis, to avoid
contamination about synonymity.
Data from a multiple sequence alignment of subtype-B HIV sequences from the LANL
website (filtered sequences only, version 2011)~\cite{LANL2012}.}
\label{fig:syndiv_genome}
\end{center}
\end{figure}

\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Alternative models for suppression of nonsynonymous mutations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.5\linewidth]{simulations_graduallyepitopesandtimeselec}
\caption{Time-dependent selection (the escape mutant is recognized after some
months from its appearance) and competition between escape mutations in the same
epitope both reduce fixation of nonsynonymous mutations, without affecting the
synonymous curve much. In the
curves denoted with ``tds'' (time-dependent selection), the fitness advantage of
an escape mutaition can vanish at each time point with a probability density
\[ P_\text{recognized}(t) = c \cdot \nu(t), \]
where $c$ is a constant coefficient shown in the legend that encodes the overall
efficiency of the host immune system, and $\nu(t)$ is the frequency of the
escape allele at time $t$. The curves denoted with ``mut/epi'' describe the
competition scenario instead, with 3 or 6 escape mutations within the same
epitope of length 9 nucleotides.}
\label{fig:tds_wec}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{natbib}
\bibliography{bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Synonymous diversity across the HIV genome}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=\linewidth]{conservation_codons_genome}
\caption{Synonymous diversity across the HIV genome, as quantified by the
normalized codon entropy. The synonymous diversity peaks at the variable
regions in {\it env} and is reduced in regions under purifying selection (RRE
hairpin, second {\it tat}/{\it rev} exons). Since the variable regions are under
positive selection at the protein level, synonymous hitchhiking is stronger
there, and even deleterious synonymous mutations can be brought to high
frequency. This gives to our signal in the fixation probability, Figs. 2A and
2B of the main text.
The normalized codon entropy is calculated as follows (see the script
for the full algorithm): (i) from a subtype B multiple sequence alignment
(MSA) from the LANL website (filtered sequences only, version 2011)~\cite{LANL2012},
we calculate the consensus amino acid at each position in the HIV genome; (ii) we count
how often each codon coding for the consensus amino acid appears in the MSA;
(iii) at each amino acid position, we divide by the number of sequences in the
MSA that had the consensus amino acid at that position, obtaining {\it codon
frequencies} $\nu_c$; (iv) we calculate the codon entropy from each position as:
$S := - \sum_{c \in \Sigma} \nu_c \log \nu_c$, where $\Sigma$ is the codon
alphabet for that particular amino acid; (v) we divide by the maximal codon entropy of
that amino acid (e.g. $\log 2$ for twofold degenerate codons).
All parts of {\it env} that are part of a different gene (signaling peptide,
second {\it rev} exon) have been excluded from our main analysis, to avoid
contamination about synonymity.}
\label{fig:syndiv_genome}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Nonsynonymous changes outside of variable regions are deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.4\linewidth]{synmut_conservation_4fold_synnonsyn}
\caption{Cumulative distribution of synonymous and nonsynonymous diversity in
{\it gag} in the LANL reference panel (filtered
sequences only, version 2011)~\cite{LANL2012}. Sites such that the consensus codon
has three synonymous and six nonsynonymous single mutants were used, and the number of observed mutants
of a certain type over the number of possible mutants is plotted. Nonsynonymous
changes are observed less often, {\it ergo} are more conserved, than
synonymous changes. It can be therefore assumed, as mentioned in the main text,
that non-escape nonsynonymous changes involve a large fitness cost.}
\label{fig:synnonsyncons}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Alternative models for suppression of nonsynonymous mutations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.5\linewidth]{simulations_graduallyepitopesandtimeselec}
\caption{Time-dependent selection (the escape mutant is recognized after some
months from its appearance) and competition between escape mutations in the same
epitope both reduce fixation of nonsynonymous mutations. A ``na\"ive'' computer
simulation shows a depression in synonymous fixation probability, as observed in
patients, but the nonsynonymous fixation probability is much higher than the
(neutral) diagonal: a nonsynonymous mutation that reaches a frequency around
0.2 is already very likely to fix. In data from infected patients, however,
nonsynonymous mutations show a neutral-like curve,
$P_\text{fix}(\nu) \approx \nu$. In the other curves, we show that the
nonsynonymous fixation probability can be reduced by either time-dependent
selection or competition between equivalent escapes, without affecting much the
synonymous fixation probability itself.
In the curves denoted with ``tds'' (time-dependent selection), the fitness advantage of
an escape mutaition can vanish at each time point with a probability density
\[ P_\text{recognized}(t) = c \cdot \nu(t), \]
where $c$ is a constant coefficient shown in the legend that encodes the overall
efficiency of the host immune system, and $\nu(t)$ is the frequency of the
escape allele at time $t$.
In the curves denoted with ``mut/epi'' we test the
competition scenario instead, with 3 or 6 escape mutations within the same
epitope of length 9 nucleotides. The fitness landscape of each such epitope
include negative epistatic terms, so that the joint presence of more than one
escape mutation is not any more beneficial for the virus than a single mutation.
In this latter scenario, many escape mutations start to sweep on different
backgrounds within the viral population, but eventually compete and only one of
them if fixed.}
\label{fig:tds_wec}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{natbib}
\bibliography{bib}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2013-02-11 synmut.tex
new Fig 4B plus some editing

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the adaptive immune system. Synonymous mutations do not affect 
the interaction with the adaptive immune system and are often
assumed to be neutral.
We analyze longitudinal intrapatient data from the \shankaregion~part of the
envelope gene (\env) and observe that synonymous derived alleles rarely
fix even though they often reach high frequencies in the viral population.
We find that synonymous mutations that disrupt base pairs in RNA stems flanking
the variable loops of gp120 are more likely to be lost than other synonymous
changes, hinting at a direct fitness effect of these stem-loop structures in the
HIV RNA.
Computational modeling indicates that these synonymous mutations have a
(Malthusian) selection coefficient of the order of $-0.002$, and that they are brought up to high
frequency by hitchhiking on neighboring beneficial nonsynonymous alleles.
The patterns of fixation of nonsynonymous mutations suggest that antibody
escape mutations in \shankaregion~are only transiently beneficial, either since
the immune system is catching up or because of competition between equivalent
escapes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are  secondary structures in the viral RNA whose disruption
is deleterious to the virus.
It has been suggested early on that parts of the viral genome that have the
potential to form RNA stems are better conserved than the
remainder~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}.

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA}, the reactivities of fixed
alleles (red histogram) are systematically larger than of alleles are lost (blue
) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to break RNA helices are also more
likely to revert and finally be lost from the population. The
average over all mutations that are not observed (green) lies between the
those that fix and those that get lost.
To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
conserved streches between the variable loops, we consider mutations in
variable loops and conserved parts separately. The biggest depression in
fixation probability is observed in the conserved stems, while the variable
loops show little deviation from the neutral signature, see \FIG{SHAPEB}. This
is consistent with important stem structures in conservered regions between loops.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $P_\text{fix}$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous HIV-like
fixation probability and diversity. Parameters are chosen uniformly on a
logarithmic scale from the indicated windows. \comment{Can't we show this
for simple epitopes? I can't seem to find a way to describe this from the
outset}. \comment{FIXME: (i) only x acis is logarithmic; (ii) redo without
epitopes (less representative in the end, but easier to introduce).}
This simulation assumes 6 roughly equivalent escape mutations per epitope; see supplement for alternative models.}
\label{fig:simheat}
\end{center}
\end{figure}

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we performed computer simulations of
evolving viral populations assuming a mix of positive and purifying selection
and rare recombination.
For this purpose, we use the simulation package FFPopSim, which includes a
module dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. 
For each simulation run, we specify the deleterious effect of synonymous
mutations, the fraction of synonymous mutations that are deleterious, the
escape rate of adaptive non-synonymous mutations and the frequency of new
escapes. Note that the escape rate is sum of two factors: (i) the beneficial
effect due to the ability to evade the immune system minus (ii) the fitness
cost of the mutation in terms of structure, stability, etc. Net escape rates
in chronic infections have been estimated to be on the order of 
$s_a = 0.01$ per day~\citep{neher_recombination_2010,Asquith}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sequence data collection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Longitudinal intrapatient viral RNA sequences were collected from published
studies~\citep{shankarappa_consistent_1999, liu_selection_2006,
bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
(LANL) HIV sequence database~\citep{LANL2012}. The samples from some patients
show substantial population structure and were discarded (see supplementary
materials); a total of 11 patients with 4-23 time points each and approximately 10
sequences per time point were analyzed. The time intervals between two
consecutive sequences ranged from 1 to 34 months, most of them approximately 6 to 10 months.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sequence analysis}
% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The sequences were translated and the resulting amino-acid sequences aligned
using Muscle~\citep{edgar_muscle:_2004} to each other and the NL4-3 reference
sequences separately for each patient. Within each patient, the consensus
nucleotide sequence at the first time point was used to classify alleles as
``ancestral'' or ``derived'' at all sites. Sites that include large
frequencies of gaps were excluded from the analysis to avoid artefactual
substitutions due to alignment errors. Allele frequencies at different time
points were extracted from the multiple sequence alignment.

A mutation was considered synonymous if it did not change the amino acid
corresponding to the codon, and if the rest of the codon was in the ancestral
state. Codons with more than one mutation were discarded. Slightly different
criteria for synonymous/nonsynonymous discrimination yield similar results.

@@

The SHAPE scores quantifying the degree of base pairing of individuals sites in
the HIV genome were downloaded from the journal
website~\citep{watts_architecture_2009}.
Where possible, SHAPE reactivities were assigned to sites in the multiple
sequence alignments for each patient through the alignment to the sequence
of the NL4.3 virus used in \citep{watts_architecture_2009}. Problematic
assignments in indel-rich regions were excluded from the analysis.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using
FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
\env{}-only simulations were performed and yielded comparable results. In each
simulation, four random parameters were drawn form prior distributions: (i)
fraction of deleterious synonymous sites (flat between 0.75 and 1.0); (ii)
deleterious effect at those sites (logarithmic between $10^{-4}$ and $10^{-2}$);
(iii) average beneficial effect of an escape mutation (logarithmic between
$10^{-2.5}$ and $10^{-1.5}$); (iv) rate of introduction of new epitopes
(logarithmic between $10^{-3}$ and $10^{-2}$ per generation). After these four
parameters were fixed, the simulation started. All the deleterious synonymous
sites had the same fitness effect (variations were tested and produced the same
results), and no epitopes (i.e. no escape mutations) were initially present.
After 30 generations of burn-in to create some genetic diversity, new epitopes
were introduced at a constant rate (as specified above). Each time such an
epitope was actually introduced, its fitness advantage was drawn from an
exponential distribution with fixed average value (as specified above).
For the models with competition within epitopes, a complex epistatic fitness
landscape was manually crafted, in which each single mutant provides the whole
beneficial effect of escape. For the models with time-dependent selection due to
the immune system catching up, the beneficial effect of an escape mutation was
reverted at random after it had appeared, according to its integrated allele
frequency (see supplementary materials for more details).

>>>>

Intrapatient HIV evolution is dominated by selection on the protein level in the
arms race with the adaptive immune system. Synonymous mutations do not affect 
the interaction with the adaptive immune system and are often
assumed to be neutral.
We analyze longitudinal intrapatient data from the \shankaregion{} part of the
envelope gene (\env{}) and observe that synonymous derived alleles rarely
fix even though they often reach high frequencies in the viral population.
We find that synonymous mutations that disrupt base pairs in RNA stems flanking
the variable loops of gp120 are more likely to be lost than other synonymous
changes, hinting at a direct fitness effect of these stem-loop structures in the
HIV RNA.
Computational modeling indicates that these synonymous mutations have a
(Malthusian) selection coefficient of the order of $-0.002$, and that they are brought up to high
frequency by hitchhiking on neighboring beneficial nonsynonymous alleles.
The patterns of fixation of nonsynonymous mutations suggest that antibody
escape mutations in \shankaregion{} are only transiently beneficial, either
because the immune system is catching up or because of competition between equivalent
escapes.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are secondary structures in the viral RNA, the disruption of
which is deleterious to the virus.
It has been suggested early on that parts of the viral genome that have the
potential to form RNA stems are better conserved than the
remainder~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}.

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA}, the reactivities of fixed
alleles (red histogram) are systematically larger than of alleles are lost (blue
) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to break RNA helices are also more
likely to revert and finally be lost from the population. The
average over all mutations that are not observed (green) lies between the
those that fix and those that get lost.
Note that, because of small differences in base pairing between the reference
strain NL4-3 and each patient's initial consensus sequence, this analysis is not
expected to be very sensitive, but very conservative.

To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
conserved streches between the variable loops, we consider mutations in
variable loops and conserved parts separately. The biggest depression in
fixation probability is observed in the conserved stems, while the variable
loops show little deviation from the neutral signature, see \FIG{SHAPEB}. This
is consistent with important stem structures in conservered regions between loops.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $P_\text{fix}$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous the
fixation probability and diversity observed in patients. Parameters are chosen
from uninformed prior distributions in the indicated window (see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we performed computer simulations of
evolving viral populations assuming a mix of positive and purifying selection
and rare recombination.
For this purpose, we use the simulation package FFPopSim, which includes a
module dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. 
For each simulation run, we specify the deleterious effect of synonymous
mutations, the fraction of synonymous mutations that are deleterious, the
escape rate of adaptive non-synonymous mutations and the frequency of new
escapes. Note that the escape rate is sum of two factors: (i) the beneficial
effect due to the ability to evade the immune system minus (ii) the fitness
cost of the mutation in terms of structure, stability, etc. Net escape rates
in chronic infections have been estimated to be on the order of 
$s_a = 0.01$ per day~\citep{neher_recombination_2010, Asquith}.
\comment{Asquith??}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Methods}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sequence data collection}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Longitudinal intrapatient viral RNA sequences were collected from published
studies~\citep{shankarappa_consistent_1999, liu_selection_2006,
bunnik_autologous_2008} and downloaded from the Los Alamos National Laboratory
(LANL) HIV sequence database~\citep{LANL2012}. The samples from some patients
show substantial population structure and were discarded (see supplementary
materials); a total of 11 patients with 4-23 time points each and approximately 10
sequences per time point were analyzed. The time intervals between two
consecutive sequences ranged from 1 to 34 months, most of them between 6 and 10
months.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Sequence analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The sequences were translated and the resulting amino-acid sequences aligned
using Muscle~\citep{edgar_muscle:_2004} to each other and the NL4-3 reference
sequences separately for each patient. Within each patient, the consensus
nucleotide sequence at the first time point was used to classify alleles as
``ancestral'' or ``derived'' at all sites. Sites that include large
frequencies of gaps were excluded from the analysis to avoid artefactual
substitutions due to alignment errors. Allele frequencies at different time
points were extracted from the multiple sequence alignment.

A mutation was considered synonymous if it did not change the amino acid
corresponding to the codon, and if the rest of the codon was in the ancestral
state. Codons with more than one mutation were discarded. Slightly different
criteria for synonymous/nonsynonymous discrimination yielded similar results.

@@

The SHAPE scores quantifying the degree of base pairing of individuals sites in
the HIV genome were downloaded from the journal
website~\citep{watts_architecture_2009}.
Where possible, SHAPE reactivities were assigned to sites in the multiple
sequence alignments for each patient through the alignment to the sequence
of the NL4.3 virus used in ref.~\citep{watts_architecture_2009}. Problematic
assignments in indel-rich regions were excluded from the analysis.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computer simulations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Computer simulations were performed using
FFPopSim~\citep{zanini_ffpopsim:_2012}. Both full-length HIV genomes and
\env{}-only simulations were performed and yielded comparable results. In each
simulation, four random parameters were drawn form prior distributions: (i)
fraction of deleterious synonymous sites (flat between 0.75 and 1.0); (ii)
deleterious effect at those sites (exponential between $10^{-4}$ and $10^{-2}$);
(iii) average beneficial effect of an escape mutation (exponential between
$10^{-2.5}$ and $10^{-1.5}$); (iv) rate of introduction of new epitopes
(exponential between $10^{-3}$ and $10^{-2}$ per generation). After these four
parameters were fixed, the simulation started. All the deleterious synonymous
sites had the same fitness effect (variations were tested and produced the same
results), no epitopes (i.e. no escape mutations) were initially present, and the
population was isomorphic.
After 30 generations of burn-in to create genetic diversity, new epitopes
were introduced at a constant rate (as specified above). Each time such an
epitope was actually introduced, its fitness advantage was drawn from an
exponential distribution with fixed average value (as specified above).
For the models with competition within epitopes, a complex epistatic fitness
landscape was manually crafted, in which each single mutant provides the whole
beneficial effect of escape. For the models with time-dependent selection due to
the immune system catching up, the beneficial effect of an escape mutation was
reverted at random times after it had appeared, according to its integrated allele
frequency (see supplementary materials for more details).
